---
title: "READ ME"
author: "Aaron Mamula"
date: "2024-06-08"
output: github_document
---

This is where you will find all the necessary information for the HPP R Tutorials.

# Purpose

The purpose of this short course is to help you develop the skills necessary to independently conduct simple data analysis in R. My hope for this course is that, after six lessons, you will be able to

* identify a dataset you would like to work with
* import those data into an R workspace
* write a simple script to extract some insight from your data

# Agenda

```{r echo=F,warning=F, message=F}
library(tidyr)
library(readxl)
library(here)
library(dplyr)
library(kableExtra)
```


```{r echo=F}
agenda <- data.frame(Date=c('7-11-2024'," "," "),
                     File=c('Lesson1'," "," "),
                     Theme=c('Introduction to R'," "," "),
                     Topics=c("Download & Install Software","Run Interactive Code",
                              "Import Data From Spreadsheet")
                     )

#agenda <- read_excel(here('HPP-R-Course-Agenda.xlsx')) %>% select(-Resources,-Code) %>% #mutate(LessonDate=format(LessonDate,"%m-%d")) %>%
#            replace(is.na(.),"")
#knitr::kable(agenda) 
knitr::kable(agenda)
```

# Lessons

The individual lessons are contained in the [Lessons Directory](https://github.com/aaronmams/HPP-R-Group-2023/tree/master/Lessons). Each lesson includes a source file with a `.Rmd` extension and a web-readable file with an `.html` extension. It will probably be easiest for you to download the `.html` file in order to follow along with the lessons.

The table above shows which `.html` files correspond to each meeting. For example, the

[Lesson1-7-13-2023.html](https://github.com/aaronmams/HPP-R-Group-2023/blob/master/Lessons/Lesson1-7-13-2023.html)

file, contains the material that will be discussed in the first meeting on 7-13-2023.



# Dependencies

Some lessons rely on external data files such as `.csv` files. These have been included in the [Data Directory](https://github.com/aaronmams/HPP-R-Group-2023/tree/master/Data). 

In the weekly meetings we will discuss the various ways you can interact with these data files. 





