---
title: "Lesson2.1-7-20-2023"
author: "Aaron Mamula"
date: "2023-06-28"
output: html_document
---

# Libraries and Packages

## How to Install Packages

From a programming and software development standpoint I think it is frowned upon to introduce this topic this early in the tutorial. From a scientific inquiry and data analysis standpoint I think this topic is critical and needs to be addressed sooner rather than later.

The process for installing packages is pretty straightforward. Normally one can just type:

```{r, eval=F}
install.packages("packagename")

```

If you copy that line into your R console it will probably tell you there is no package called "package name" but you may be able to execute this line:

```{r, eval=F}
install.packages("dplyr")

```

Once a package is installed, it still must be loaded into the active workspace. This can be done as follows:

```{r}
require(dplyr)

```

Sometimes, weird things happen and packages don't install. Sometimes it can be beneficial to try the point-and-click way:

* Packages --> install packages if you are using the R Gui
* Tools --> install packages in RStudio


Sometimes, installing and using packages gets funky if you don't understand exactly where R is installing and looking for external packages. In such cases, a good first step is to look at where R is looking for packages:

```{r}
.libPaths()

```

This will give you the file locations where R is looking to find packages when you do a ```require(packagename)```. 

In cases where the place that R is installing packages when you do ```install.packages(packagename)``` is not the same place that R is looking for packages when you do ```require(packagename)```, the solution to overcome this will be a little difficult. 

[I suggest starting here](https://www.accelebrate.com/library/how-to-articles/r-rstudio-library). I am also available to help troubleshoot these types of issues.


## Why Do I Need to Install Packages

I encourage you to do some of your own reading and form your own opinions about how much you want to rely on external packages versus base code. The R approach of having a lot functionality "baked in" to base code then allowing lots of other functionalities to be supported by other open source libraries has the following flexibility:

it makes R a very flexible language while still being a relatively easy to use language. In the world of computers and code there will always be a tradeoff between scope (the universe of things you can do) and approachability (how easy it is to do things).

C is a language with a lot of flexibility. You can do anything from complex biostatistics to building an iPhone app in C. However, doing anything C requires substantial expertise in both the subject matter and the programming language. You *can* take the mean of a group of numbers in C but you can't tell C something like:

```{r}
mean(2,3,4)

```

Instead you would do something like,

```{r, eval=F}
int i, j, k; 
double meanval;

i=10;
j=20;
k=30;

meanval = (i+j+k)/3

```

On the other hand, Excel makes it very easy to take the mean value of some numbers. You just point-and-click the *average* function and highlight the cells you want to use.

To recap:

* C has a lot of flexibility but also a lot of complexity. In C you are not limited only to the things C says you can do, you are only limited by the things you are able to do.
* Excel has no flexibility but also very little complexity. In Excel you are completely limited only to operations that Excel approves of - you can take the mean of some number, you can make a bar chart, but you will never be able to program an iPhone app.

R's approach to navigate this tradeoff between flexibility and complexity is to offer some things you can do with base code - make graphs, add numbers, store data - and leave other functionalities to be developed by the users. The result is a very powerful piece of software that can do highly complex analysis across a wide range of disciplines (finance, political science, biology, hydrology, oceanography, physics, psychology, etc.) but is still relatively easy to use. 


 


